Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models

### 2015-04-09 tex/parametrized-learning.tex
first pass at new introduction

In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
Directly constructing the likelihood ratio for high-dimensional observations 
is often not possible or is computationally impractical. Here we demonstrate how 
discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for training and calibration.  We use the following notation

\begin{itemize}
 \item $x$: a vector of features
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $p(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
\item $p( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.

\newpage

In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio 
\begin{equation}
T(D) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
\end{equation}
is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density 
$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
high energy physics where the simulation of particle detectors can only be done in the `forward mode'.

Our main result is that one can form an equivalent test based on 
%\begin{equation}
%T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
%\end{equation}
\begin{equation}\label{eq:equivLRtest}
T'(D) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
\end{equation}
if 
\begin{equation}\label{eq:montonic}
s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \; 
%s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \; 
\end{equation}
where $m$ is any strictly increasing or decreasing function. This result will be proven below.
This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$. In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.

@@

\section{Comments on classification and frequentist hypothesis tests}

@@

\section{Composite hypotheses and the generalized likelihood ratio}\label{?}l{S:GLR}

>>>>

In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
Directly constructing the likelihood ratio for high-dimensional observations 
is often not possible or is computationally impractical. Here we demonstrate how 
discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for training and calibration.

As a concrete example, consider searches for new particles at the Large Hadron Collider. 
The ATLAS and CMS experiments have published several hundred papers where the 
final statistical result was formulated as a hypothesis test or confidence interval 
using a generalized likelihood ratio test~\citep{Cowan:2010js}. This includes
the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and 
subsequent measurement of its properties.  
The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
that discriminates between a process of interest (labeled \textit{signal}) and various other processes 
(labeled \textit{background}).

To improve the statistical power of these tests, hundreds of these searches have utilized discriminative
classifiers trained on labeled data. Typically using libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht}, though recently there has been progress in deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history. The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.

The labeled data used to train the classifiers are not obtained from the experimental apparatus. 
Instead they are generated from a 
detailed simulation, which includes a quantum mechanical description of the underlying particle 
interactions, an exquisitely detailed simulation of the particle detectors with hundreds of millions of 
sensors, and the data processing pipeline that reduces the data for a single event to tens or hundreds of high-level features used as input to the classifier. 
It takes several minutes for this generative model to simulate a single collision or \textit{event} with Monte Carlo
techniques. While particle physics offers an extreme example, it is common in science that generative 
models are available while it not feasible to evaluate the probability density function that is implicitly being 
sampled by the generative model.

This paper describes how one can utilize generative statistical models for the data in conjunction with 
discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to 
the one that would have been obtained if it were possible to evaluate the probability 
density implicitly being sampled by the generative model.

\subsection{Notation and Assumptions}

We use the following ntoootation:
\begin{itemize}
 \item $x$: a vector of features
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $p(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.

%\newpage
\subsection{Prelude}

In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio 
\begin{equation}
T(D) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
\end{equation}
is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density 
$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \cite{ClaytonScott} and \cite{JMLR:v14:tong13a}.

The main result of this paper is that one can form an equivalent test based on 
%\begin{equation}
%T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
%\end{equation}
\begin{equation}\label{eq:equivLRtest}
T'(D) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
\end{equation}
if 
\begin{equation}\label{eq:montonic}
s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \; 
%s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \; 
\end{equation}
where $m$ is any strictly increasing or decreasing function. This result will be proven below.
This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$. In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.

@@

\subsection{Comments on classification and frequentist hypothesis tests}

@@

\section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
